Sampling-Based Motion Planning for Tracking Evolution of Dynamic Tunnels in Molecular Dynamics Simulations

### 2017-10-31 paper.tex
BK - rw checked

The analysis of protein structure aiming to reveal tunnels has been supported by different computational software tools, which take the geometry of the protein as an input and explore the inner void space (e.g., MOLE~\cite{Petrek20071357} or CAVER 1.0~\cite{petrek2006caver}).
Early methods for tunnel detection utilized a discretized 3D grid, where each cell was considered as occupied or free depending on the presence of atoms of the protein.
Tunnels can be then searched using standard graph-search methods, such as the Dijkstra's algorithm.
Besides, the grid can be used to identify other relevant properties like pockets, cavities, or channels~\cite{sehnal2013mole,petrek2006caver}.
%One of the first grid-based approaches to the detection of tunnels in protein is the CAVER 1.0 algorithm by Pet\v{r}ek et al.~\cite{citeulike:6257975}.
The obvious disadvantage of the grid-based methods is the high memory demand and their dependency on the grid resolution.
Due to the high memory consumption, these methods are not suitable for tunnel detection in dynamic proteins and therefore they are used primarily for the analysis of static molecules or individual snapshots of the dynamics.

Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi diagrams (WVD).
The ordinary VD is computed on points representing centers of all atoms, without considering the atomic radii.
This may lead to the detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
To consider atoms with different radii, the weights of individual points are determined by the van der Walls radii of the atoms in WVD.
An alternative solution is to compute a non-weighted VD on an extended point set, where each atom is approximated by several spheres with a small radius~\cite{yaffe2008,caver3,caverDetails}.
VD-based methods are memory less demanding, and also faster than the grid-based methods.
The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and finding the correspondences between them.
The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
Another problem of these approaches is the definition of tunnel bottleneck threshold, which determines the minimal size of the narrowest tunnel point.
If the bottleneck radius is even slightly below this threshold, the tunnel is discarded. 
This leads to the problem of missing tunnels in several or many snapshots of the MD simulation, even when the void space is still present in them.

Alternatively, the tunnel detection in dynamic proteins can be formulated as path planning in 4D (i.e., 3D $\times$ time) configuration space assuming a spherical probe.
This space can be searched using sampling-based motion planning methods like Probabilistic Roadmaps (PRM)~\cite{kavrakiPRM} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} and their modifications.
In sampling-based motion planning, the configuration space is randomly sampled and the samples are classified as free or non-free using collision detection.
The collision detection is performed between the geometry of the robot (i.e., the probe) and the environment (the atoms of the protein).
The collision-free samples are stored into a roadmap which is searched using standard graph-search methods for a path.
The path in the roadmap then describes a motion in the workspace.

The advantage of sampling-based motion planners is that they can cope with many-DOF (Degrees Of Freedom) robots and due to black-box collision detection, they can also cope with robots of arbitrary shapes.
Beyond robotics, the sampling-based motion planners have been successfully applied to computational biology problems in studies related to protein folding~\cite{bayazit2001ligand,al2012motion,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro,kirillova2008an}, ligand docking~\cite{cortes2010simulating,latombe1999motion,mollProt}, or analysis of loop motions~\cite{cortes2004geometric}.
In these applications, the molecules are represented as a kinematic chain where joint angles correspond to bonds of the molecules.

%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
RRT is a single-query motion planning method.
It incrementally builds a configuration tree $\T$ rooted at the initial configuration.
In each iteration of RRT, a random configuration $\qrand$ is generated and its nearest node $\qnear$ in the tree  is found.
A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ (straight-line expansion).
If $\qnew$ is collision-free, it is added to the tree.
The algorithm terminates if the tree approaches the goal configuration to a predefined distance.

In~\cite{guieysse2008structure,cortes2010simulating}, RRT is used to compute exit pathways for a flexible ligand leading from an active site to the surface of the protein.
To cope with many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
In~\cite{cortesMD}, the pathway is found for the opposite direction, from the active site towards the protein surface.
The methods~\cite{guieysse2008structure,cortesMD} do not compute all tunnels explicitly, rather a single pathway.

@@

A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
in the difficult areas.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling, e.g., around medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}, or even by shifting uniformly generated samples towards the medial axis~\cite{amatoOBPRM} or its approaximation~\cite{hollemanMAPRM}.
Another approach is to estimate difficult regions online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
Generating random samples in difficult regions however does not ensure the construction of a better tree in the case of RRT-based methods.
As RRT builds the tree incrementally, the tree can connect to a region with increased probability of sampling only if the growth of the tree is not blocked by obstacles.
Instead of increasing probability of sampling in all regions at once, e.g., along the whole medial axis, the tree has to be guided through the space to reach the regions gradually.
For example, the random samples are generated from multiple regions, that are switched according to the progress of the tree.
In~\cite{kardossRRTKK}, the samples are generated along several waypoints that are located close to narrow passages.
A more general approach is to define a guiding path in the workspace and sequentially increase the probability of sampling along the path~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}.
%The guiding path can be computed e.g. using medial axis of the worskpace~\cite{denny2014marrt,denny2016dynamic}.

>>>>

The analysis of protein structure aiming to reveal tunnels has been supported by different computational software tools, which take the geometry of the protein as an input and explore the inner void space (e.g., MOLE~\cite{Petrek20071357} or CAVER 1.0~\cite{petrek2006caver}).
Early methods for tunnel detection utilized a discretized 3D grid, where each cell was considered as occupied or free, depending on the presence of atoms of the protein.
Tunnels can be then searched using standard graph-search methods, such as the Dijkstra's algorithm.
Besides, the grid can be used to identify other relevant properties, like pockets, cavities, or channels~\cite{sehnal2013mole,petrek2006caver}.
%One of the first grid-based approaches to the detection of tunnels in protein is the CAVER 1.0 algorithm by Pet\v{r}ek et al.~\cite{citeulike:6257975}.
The obvious disadvantage of the grid-based methods is the high memory demand and their dependency on the grid resolution.
Namely due to the high memory consumption, these methods are not suitable for tunnel detection in molecular dynamics and therefore they are used primarily for the analysis of static molecules or individual snapshots of the molecular dynamics.

Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi diagrams (WVD).
The ordinary VD is computed on points representing centers of all atoms, without considering the atomic radii.
This may lead to the detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
To overcome this, the weights of individual points can be determined by the van der Walls radii of the atoms in WVD.
An alternative solution is to compute a non-weighted VD on an extended point set, where each atom is approximated by several spheres with a same small radius~\cite{yaffe2008,caver3,caverDetails}.
VD-based methods are memory less demanding, and also faster than the grid-based methods.
The extension of the VD-based methods to molecular dynamics requires to construct the VD in all frames of the simulation and finding the correspondence over time between the tunnels detected in these frames.
The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
Another problem of these approaches is the definition of the tunnel bottleneck threshold, which determines the minimal size of the narrowest tunnel point.
If the bottleneck radius is even slightly below this threshold, the tunnel is discarded. 
This leads to the problem of missing tunnels in several or many snapshots of the MD simulation, even when the void space is still present in them.

Alternatively, the tunnel detection in dynamic proteins can be formulated as a path planning problem in 4D (i.e., 3D $\times$ time) configuration space, assuming a spherical probe.
This space can be searched using sampling-based motion planning methods, like Probabilistic Roadmaps (PRM)~\cite{kavrakiPRM} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} and their modifications.
In the sampling-based motion planning, the configuration space is randomly sampled and the samples are classified as free or occupied using collision detection.
The collision detection is performed between the geometry of the robot (i.e., the probe) and the environment (the atoms of the protein).
The collision-free samples are stored into a roadmap which is searched using standard graph-search methods for a path.
The path in the roadmap then describes a motion in the workspace.

The advantage of sampling-based motion planners is that they can cope with many-DOF (Degrees Of Freedom) robots and due to the black-box collision detection, they can also cope with robots of arbitrary shapes.
Beyond robotics, the sampling-based motion planners have been successfully applied to computational biology problems in studies related to protein folding~\cite{bayazit2001ligand,al2012motion,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro,kirillova2008an}, ligand docking~\cite{cortes2010simulating,latombe1999motion,mollProt}, or analysis of loop motions~\cite{cortes2004geometric}.
In these applications, the molecules are represented as a kinematic chain where the joint angles correspond to bonds of the molecules.

%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
RRT is a single-query motion planning method.
It incrementally builds a configuration tree $\T$ rooted at the initial configuration.
In each iteration of the RRT, a random configuration $\qrand$ is generated and its nearest node $\qnear$ in the tree  is found.
A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ (straight-line expansion).
If $\qnew$ is collision-free, it is added to the tree.
The algorithm terminates if the tree approaches the goal configuration to a predefined distance.

In~\cite{guieysse2008structure,cortes2010simulating}, the RRT is used to compute exit pathways for a flexible ligand leading from the active site to the surface of the protein.
To cope with many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
The RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
In~\cite{cortesMD}, the pathway is found for the opposite direction, from the active site towards the protein surface.
The methods~\cite{guieysse2008structure,cortesMD} do not compute all tunnels explicitly, rather a single path.

@@

A possible solution is to estimate the location of the narrow passages from the knowledge of the workspace and generate more samples in the difficult areas.
The PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling, e.g., around the medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}, or even by shifting the uniformly generated samples towards the medial axis~\cite{amatoOBPRM} or its approximation~\cite{hollemanMAPRM}.
Another possible approach is to estimate the difficult regions online, e.g., based on the number of the collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
However, generating random samples in the difficult regions does not ensure the construction of a better tree in the case of RRT-based methods.
As the RRT builds the tree incrementally, the tree can connect to a region with increased probability of sampling only if the growth of the tree is not blocked by obstacles.
Instead of increasing the probability of sampling in all regions at once, e.g., along the whole medial axis, the tree has to be guided through the space to reach the regions gradually.
For example, the random samples are generated from multiple regions, that are switched according to the progress of the tree.
In~\cite{kardossRRTKK}, the samples are generated along several waypoints that are located close to the narrow passages.
A more general approach is to define a guiding path in the workspace and sequentially increase the probability of sampling along the path~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}.
%The guiding path can be computed e.g. using medial axis of the worskpace~\cite{denny2014marrt,denny2016dynamic}.
